pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-07-03 bmc_article.tex

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

@@

{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}. Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space. However, these studies, like ours, used datasets with relatively few cells (i.e. fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted. Further work will therefore be required to assess compare its performance on larger datasets.}

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines. In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts. We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.

@@

{\color{red}\subsection*{Overall run for interactions across steps}

@@

Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument. For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2} \citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}. \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit. The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

@@

{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}. Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space. These studies, like ours, used datasets with relatively few cells (i.e. fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted. We think that further work will be required to assess its performance on larger datasets.}

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different workflow from the other pipelines. In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts. We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.

@@

\subsection*{Overall run for interactions across steps}

@@

Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument. For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2} \citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}. \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit. The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
